A Logical Solution of the Linearity Challenge

### 2025-11-29 paper/calculus.tex
revised description of calculus

The types for \Calculus, ranged over by $A$, $B$, \dots, are linear logic
propositions generated by the grammar
\[
    A, B ::= X \mid \dual{X} \mid \Zero \mid \Top \mid \One \mid \Bot \mid A \Plus B \mid A \With B \mid A \Ten B \mid A \Par B \mid \OfCourse A \mid \WhyNot A \mid \exists X.A \mid \forall X.A
\]
where $X$, $Y$, \dots range over an infinite set of \emph{type variables}.

The interpretation of propositions as session types is the usual
one~\cite{Wadler14}: the constants $\Bot$ and $\One$ describe the input/output
of a termination signal; the additive connectives $A \With B$ and $\A \Plus B$
describe the input/output of a label $\InTag_i$ followed by the corresponding
protocol $A$ and $B$; the multiplicative connectives $A \Par B$ and $A \Ten B$
describe the input/output of a channel (to be used according to $A$) followed by
the protocol $B$. The additive constants $\Zero$ and $\Top$ do not describe
useful protocols, but they can serve the role of smallest/largest type in type
systems that support a notion of subtyping~\citep{HornePadovani24}. The ``of
course'' modality $\OfCourse A$ and the ``why not'' modality $\WhyNot A$
describe servers and clients accepting and requesting connections to be used
according to $A$. Finally, the quantifiers $\forall X.A$ and $\exists X.A$
describe the input/output of a type $X$ followed by the protocol $A$.

The notions of \emph{duality} and of \emph{type substitution} are the usual
ones~\cite{Wadler14}. Specifically, we write $\dual{A}$ for the dual of $A$ and
$A\subst{B}X$ for the type obtained by replacing the free occurrences of $X$ in
$A$ with $B$.

@@

\begin{table}
    \caption{Syntax of \Calculus.}\label{tab:syntax}
    \begin{tabular}{c}
        \begin{math}
            \displaystyle
            \begin{array}[t]{rcll}
                P, Q & ::= & \Link\x\y            & \text{link} \\
                    & | & \Fail\x                 & \text{fail} \\
                    & | & \Wait\x.P               & \text{wait} \\
                    & | & \Close\x                & \text{close} \\
                    & | & \Case\x\z{P}{Q}         & \text{case} \\
                    & | & \Select\x{\InTag_i}\z.P & \text{selection} \\
                    & | & \Join\x\y\z.P           & \text{join} \\
                    & | & \Fork\x\y\z{P}{Q}       & \text{fork} \\
                    & | & \Forall\x\X\y.P         & \text{type input} \\
                    & | & \Exists\x{A}\y.P        & \text{type output} \\
                    & | & \Server\x\y.P           & \text{server accept} \\
                    & | & \Client\x\y.P           & \text{client request} \\
                    & | & \Weaken\x.P             & \text{weakening} \\
                    & | & \Contract\x\y\z.P       & \text{contraction} \\
                    & | & \Cut[A]\x{P}{Q}         & \text{cut} \\
            \end{array}
        \end{math}
    \end{tabular}
\end{table}

The syntax of \Calculus makes use of an infinite set of \emph{channel names},
ranged over by $x$, $y$ and $z$, and is shown in~\Cref{tab:syntax}.
%
A link $\Link\x\y$ denotes the merging of the channels $x$ and $y$, so that
every message sent on one of the channels is forwarded to the other. As
discussed in the literature, this form is useful also for modeling
\emph{delegation}, that is the exchange of an existing session endpoint in
another session.
%
The $\Close\x$ and $\Wait\x.P$ processes respectively model the output and input
of a termination signal on the channel $x$. The former process has no
continuation, while the latter process continues as $P$ once the signal has been
received.
%
The process $\Fail\x$ denotes a runtime error concerning the channel $x$.
%
The processes $\Case\x\z{Q_1}{Q_2}$ and $\Select\x{\InTag_i}\z.P$ for $i=1,2$
describe the input/output of a label $\InTag_i$ along with a fresh continuation
$z$ on the channel $x$. The receiver continues as $Q_i$ depending on the
received label $\InTag_i$. In practice it is useful to consider arbitrary
(domain-specific) sets of tags instead of the canonical $\InTag_1$ and
$\InTag_2$. This generalization has no substantial impact on the formalization,
so we stick to the classical case of ``binary choices'' in line with the usual
presentations of linear logic.
%
The processes $\Join\x\y\z.R$ and $\Fork\x\y\z{P}{Q}$ describe the input/output
of a pair of fresh channels $y$ and $z$ on the channel $x$. The former process
forks into $P$ and $Q$, each using $y$ and $z$ respectively. The latter process
continues as $R$. These forms are often used in conjunction with links to model
\emph{delegation}, but can also be used to model bifurcating sessions whereby a
sequential protocol continues as two parallel sub-protocols.
%
The processes $\Exists\x{A}\y.P$ and $\Forall\x\X\y.P$ describe the input/output
of a type on the channel $x$ along with a fresh continuation $y$. These forms
are useful to describe polymorphic processes.

@@

The notions of free and bound names are fairly standard. Just bear in mind that
also output action prefixes bind continuation channels in (some) continuation
processes. In particular, $\Select\x{\Tag_i}\z.P$ binds $z$ in $P$, while
$\Fork\x\y\z{P}{Q}$ binds $y$ in $P$ but not in $Q$ and $z$ in $Q$ but not in
$P$.
%
We write $\fn{P}$ for the set of channel names occurring free in $P$ and we
identify processes up to renaming of bound names.

@@

Notably and somewhat surprisingly, there is no pre-congruence rule asserting
that parallel composition is associative. While this is a sound and expected
property of parallel composition, it turns out to be inessential for proving all
of the properties of \Calculus that we consider. We leave it out in light of the
forthcoming formalization of the calculus, in which every simplification is
welcome, although standard presentations of classical linear logic admit such
reduction~\cite{Wadler14,Doumane17}.
%
Another rule that deserves attention is \SServer. In this case, a server prefix
$\Server\y\u$ can be pulled out of a cut only provided that the other process in
the cut is also a server on the channel restricted by the cut.

@@

The rule~\RCut propagates reductions through cuts and the rule \RCong closes
reduction by structural pre-congruence.

@@

Typing contexts (\ie sequents) are finite maps from channel names to types
written as $x_1 : A_1, \dots, x_n : A_n$ and ranged over by $\ContextC$ and
$\ContextD$. We write $\ContextC, \ContextD$ for the union of $\ContextC$ and
$\ContextD$ when they have disjoint domain. We write $\WhyNot\Context$ for some
context $\Context = x_1 : \WhyNot A_1, \dots, x_n : \WhyNot A_n$ where all the
types in its range are prefixed by the $\WhyNot$ modality. We call these types
\emph{unrestricted}, since they are used to denote shared channels that can be
(explicitly) discarded and duplicated at will.

@@

We now shift the attention to \emph{well formedness}. In the linearity
challenge~\cite{CarboneEtAl24}, this property ensures that, whenever two
processes composed in parallel begin with actions concerning the very same
session, then such actions are \emph{complementary}, in the sense that they
describe opposite forms of interaction. To define well formedness in our
setting, we introduce \emph{reduction contexts} as partial processes with a
single unguarded hole $\Hole$, thus:
\[
    \ReductionContext ::= \Hole \mid \Cut\x\ReductionContext{P} \mid \Cut\x{P}\ReductionContext
\]

>>>>

The types of \Calculus, ranged over by $A$, $B$, \dots, are the linear logic
propositions generated by the grammar
\[
    A, B ::= X \mid \dual{X} \mid \Top \mid \Zero \mid \Bot \mid \One \mid A \With B \mid A \Plus B \mid A \Par B \mid A \Ten B \mid \forall X.A \mid \exists X.A \mid \OfCourse A \mid \WhyNot A
\]
where $X$, $Y$, \dots range over an infinite set of \emph{type (or proposition)
variables}.

The interpretation of propositions as session types is the usual
one~\cite{Wadler14}: the multiplicative constants $\Bot$ and $\One$ describe
protocols that terminate with the input/output of a signal; the additive
connectives $A \With B$ and $\A \Plus B$ describe protocols having two possible
continuations $A$ and $B$ with are selected/offered; the multiplicative
connectives $A \Par B$ and $A \Ten B$ describe bifurcating protocols with two
continuations $A$ and $B$. These forms are typically used to describe the
input/output of a channel (to be used according to $A$) and then continuing as
$B$. The additive constants $\Zero$ and $\Top$ do not describe useful protocols,
but they can play the role of smallest/largest type in type systems that support
a notion of subtyping~\citep{HornePadovani24}. 
%
The quantifiers $\forall X.A$ and $\exists X.A$ describe the input/output of a
type $X$ followed by the protocol $A$.
%
Finally, the ``of course'' modality $\OfCourse A$ and the ``why not'' modality
$\WhyNot A$ describe servers and clients accepting and requesting connections to
be used according to $A$.

The notions of \emph{free type variables}, \emph{duality} and of \emph{type
substitution} are standard~\cite{Wadler14}. In particular, we write $\fv{A}$ for
the set of type variables occurring free in $A$, we write $\dual{A}$ for the
dual of $A$ and $A\subst{B}X$ for the type obtained by replacing the free
occurrences of $X$ in $A$ with $B$.

@@

\begin{table}
    \caption{Syntax of \Calculus.}\label{tab:syntax}
    \begin{tabular}{c}
        \begin{math}
            \displaystyle
            \begin{array}[t]{rcll}
                P, Q & ::= & \Link\x\y            & \text{link} \\
                    & | & \Fail\x                 & \text{empty choice} \\
                    & | & \Wait\x.P               & \text{wait} \\
                    & | & \Close\x                & \text{close} \\
                    & | & \Case\x\z{P}{Q}         & \text{case} \\
                    & | & \Select\x{\InTag_i}\z.P & \text{selection} \\
                    & | & \Join\x\y\z.P           & \text{join} \\
                    & | & \Fork\x\y\z{P}{Q}       & \text{fork} \\
                    & | & \Forall\x\X\y.P         & \text{type input} \\
                    & | & \Exists\x{A}\y.P        & \text{type output} \\
                    & | & \Server\x\y.P           & \text{server accept} \\
                    & | & \Client\x\y.P           & \text{client request} \\
                    & | & \Weaken\x.P             & \text{weakening} \\
                    & | & \Contract\x\y\z.P       & \text{contraction} \\
                    & | & \Cut[A]\x{P}{Q}         & \text{cut} \\
            \end{array}
        \end{math}
    \end{tabular}
\end{table}

The syntax of \Calculus makes use of an infinite set of \emph{channel names},
ranged over by $x$, $y$ and $z$, and is shown in~\Cref{tab:syntax}.
%
A link $\Link\x\y$ denotes the merging of the channels $x$ and $y$, so that
every message sent on one of the channels is forwarded to the other. As
discussed in the literature~\cite{LindleyMorris}, this form is also useful for
modeling \emph{delegation}, that is the exchange of an existing session endpoint
in another session.
%
The $\Wait\x.P$ and $\Close\x$ processes respectively model the input and output
of a termination signal on the channel $x$. The latter process has no
continuation, while the former process continues as $P$ once the signal has been
received.
%
The process $\Fail\x$ can be used to denote a failing protocol concerning the
channel $x$.
%
The processes $\Case\x\z{Q_1}{Q_2}$ and $\Select\x{\InTag_i}\z.P$ for $i=1,2$
describe the input/output of a label $\InTag_i$ along with a fresh continuation
$z$ on the channel $x$. The receiver continues as $Q_i$ depending on the
received label $\InTag_i$. In practice it is useful to consider arbitrary
(domain-specific) sets of tags instead of the canonical $\InTag_1$ and
$\InTag_2$. This generalization has no substantial impact on the formalization,
so we stick to the classical case of ``binary choices'' in line with the usual
presentations of linear logic.
%
The processes $\Join\x\y\z.R$ and $\Fork\x\y\z{P}{Q}$ describe the input/output
of a pair of fresh channels $y$ and $z$ on the channel $x$. The former process
forks into $P$ and $Q$, each using $y$ and $z$ respectively. The latter process
continues as $R$. These forms are often used in conjunction with links to model
\emph{delegation}.
%
The processes $\Exists\x{A}\y.P$ and $\Forall\x\X\y.P$ describe the input/output
of a type on the channel $x$ along with a fresh continuation $y$. These forms
are useful to describe polymorphic processes.

@@

The notions of free and bound names are fairly standard, bearing in mind that
output action prefixes bind continuation channels in (some) continuation
processes. In particular, $\Select\x{\Tag_i}\z.P$ binds $z$ in $P$, while
$\Fork\x\y\z{P}{Q}$ binds $y$ in $P$ but not in $Q$ and $z$ in $Q$ but not in
$P$.
%
We write $\fn{P}$ for the set of channel names occurring free in $P$ and we
identify processes up to renaming of bound names.

@@

Notably and somewhat surprisingly, there is no pre-congruence rule asserting
that parallel composition is associative. While this is a sound and expected
property of parallel composition, it turns out to be inessential for proving all
of the properties of \Calculus that we consider. We leave it out in light of the
forthcoming formalization of the calculus, in which every simplification is
welcome, although standard presentations of classical linear logic admit such
law~\cite{Wadler14,Doumane17}.
%
Another rule that deserves attention is \SServer. In this case, a server prefix
$\Server\y\u$ can be pulled out of a cut only provided that the other process in
the cut is also a server on the channel restricted by the cut.

@@

The rule~\RCut propagates reductions through cuts and the rule \RCong enables
reduction up to structural pre-congruence.

@@

Typing contexts (\ie sequents) are finite maps from channel names to types
written as $x_1 : A_1, \dots, x_n : A_n$ and ranged over by $\ContextC$ and
$\ContextD$. We write $\ContextC, \ContextD$ for the union of $\ContextC$ and
$\ContextD$ when they have disjoint domains. We write $\WhyNot\Context$ for some
context $\Context = x_1 : \WhyNot A_1, \dots, x_n : \WhyNot A_n$ where all the
types in its range are prefixed by the $\WhyNot$ modality. We call these types
\emph{unrestricted}, since they are used to denote shared channels that can be
(explicitly) discarded and duplicated at will.

@@

We now shift the attention to \emph{well formedness}. In the linearity
challenge~\cite{CarboneEtAl24}, this property ensures that, whenever two
processes composed in parallel begin with actions concerning the very same
session, then such actions are \emph{complementary}, in the sense that they
describe opposite forms of interaction. To define well formedness in our
setting, we introduce \emph{reduction contexts} as partial processes with a
single unguarded hole $\Hole$, thus:
\[
    \ReductionContext ::= \Hole \mid \Cut[A]\x\ReductionContext{P} \mid \Cut[A]\x{P}\ReductionContext
\]
